Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-05 paper.Rmd
moar editing
I said moar

---
title: "Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: Richard K. Morgan and Andreas Beger and Michael D. Ward
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_tex: true
references:
- id: blair2020forecasting
  title: "Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning"
  author:
  - family: Blair
    given: Robert A.
  - family: Sambanis
    given: Nicholas
  container-title: Journal of Conflict Resolution
  type: article-journal
  issued:
    year: 2020
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else references.bib`"
---

@@

Blair and Sambanis (2020; hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.  Indeed they argue that with such theory, forecasting is more accurate than has previously been possible.  Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article.  We find that these claims are false.  Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.  The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure: 1) they use smoothed performance curves instead of the unsmoothed curves, 2) they mistakenly encode the dependent variable to be incidents of civil war, rather than as they claim the onset of a civil war, 3) they mistakenly calculate the predicted values for the alternative "non-theoretical" models that they compare to.  In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.  We believe that this research by B&S is incorrect and should be retracted.

@@

@blair2020forecasting (B&S for the sake of brevity hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed. They find that it does. They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,000 covariates.

@@

- (1) How does the theoretically-driven escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
- (2) Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
- (3) How accurate were predictions for the first half of 2016 derived from the escalation model?

To assess the first two questions, B&S use global datasets covering all major countries from 2001 to 2015. Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years. The main outcome variable is civil war onset, measured using Sambanis' civil war dataset.

Both the first and second questions above rely on comparing their Escalation model to various alternative models. The same procedure is used in both cases:

1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
2. Estimate the Escalation and other competing models. 
3. Create out-of-sample (OOS) predictions from each model using the test set. 
4. Calculate AUC-ROC measures for each set of OOS predictions.

The corresponding results for each question are shown in B&S Tables 1 and 2, which we will replicate further below.

To examine the first question, B&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other. These are directed, thus making for four total covariates.
- CAMEO: counts for all CAMEO event codes, thus a total of 1,159 covariates (CAN THIS BE RIGHT?). 
- Average: unweighted average of the predictions from the four models briefly described above.

The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions. These robustness checks vary either (1) random forecast hyperparameter values, or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.

The second question, whether structural variables add to the Escalation model, is assessed by comparing the original escalation model to four alternatives that incorporate annual, structural variables that are used in the PITF instability forecasting model:

@@

Finally, B&S used their Escalation model to create forecasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed. This is summarized in B&S Table 3.

@@

While attempting to replicate B&S's results, we found several issues.

@@

This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.

@@

ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data. In this case, there are also groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached. For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates. We can only speculate, but maybe the very "step"-like nature of the ROC curves plus the presence of unusual diagonal lines accounts for the decision to use smoothed ROC curves.

As it turns out, this decision has a dramatic impact on the AUC-ROC calculations they use to support their empirical claims. As we will show in the results further below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and deferentially impacts the models that are being compared. In fact, B&S's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC.

@@

We infer that the intent is that the Escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. Their implementation, however, actually weighs the test set predictions using the PITF model predictions for the training set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly in the 6-month version of this file.] This appears to be a coding error.

@@

The implementation error aside, this split-population analog model is actually quite odd and does not actually replicate an analogue of the idea behind split-population modeling. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision (regression^[See further below. Although the RF models are used for a binary decision problem, the actual implementation uses regression RFs for continuous outcomes.]) trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does indeed have 200,000 decision trees. Furthermore, the PITF model predictions do not impact the way the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking just another Escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).

@@

## Basic interpretation of Table 1

Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC. The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models. It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel. In any case, the interpretation of results should not be sensitive to random variation, i.e. it should not depend on using a specific RNG seed. On the basis of these results, B&S conclude that the Escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.

Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves. The Average model outperforms the Escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.

@@

table2_nosmooth %>%
  dplyr::select(-Model, -horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of B\\&S Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution",
    label = "table2-fixed") %>%
  column_spec(2:6, width = "2cm") %>%
  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
  footnote("Differences from the original B\\&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
```

@@

This is not the case anymore. It outperforms in the 1-month version and under-performs the Escalation model in the 6-month version.

@@

What impact did the ROC smoothing have overall on the results reported in B&S Tables 1 and 2? Tables \ref{tab:table1-benefit} and \ref{tab:table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC. Positive values indicate that smoothing increased a model's apparent performance. The Escalation model is the only model that consistently had a benefit from smoothing. For all eight other models, smoothing sometimes gave a benefit, sometimes not.

A more dramatic difference stands out when we consider the overall average impact of smoothing across all flavors of a model reported in the columns of the tables. The Average, Goldstein, Quad, PITF Split Population, and Weighted by PITF models are slightly hurt by smoothing, but on the order of less than 0.01 in absolute magnitude. The "With PITF Predictors" model is hurt on the order of 0.02, i.e. it appears substantially worse with smoothing. The CAMEO and PITF Only models benefit slightly, on the order of 0.01 or less. The Escalation Only model on the other hand has an average boost of 0.058 to its AUC-ROC from smoothing. Considering the spread of AUC-ROC values if we compare across rows in B&S Tables 1 and 2, that boost is substantial.

The use of smoothed ROC curves to calculate AUC-ROC values only clearly benefits the Escalation model, it does so consistently, and by a considerable margin. All eight alternative models reported in Tables 1 and 2 do not on average gain when using smoothed ROC curves to calculate AUC.

@@

- the theory-driven Escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
- structural variables substantially improve the Escalation model's performance when added to the pool of predictors the underlying random forest model draws on

>>>>

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: Andreas Beger and Richard K. Morgan and Michael D. Ward
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_tex: true
references:
- id: blair2020forecasting
  title: "Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning"
  author:
  - family: Blair
    given: Robert A.
  - family: Sambanis
    given: Nicholas
  container-title: Journal of Conflict Resolution
  type: article-journal
  issued:
    year: 2020
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else references.bib`"
---

@@

Blair and Sambanis (2020; hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.  Indeed they argue that with such theory, forecasting is more accurate than has previously been possible.  Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article.  We find that these claims are false.  Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.  The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure: 1) they use smoothed performance curves instead of the unsmoothed curves, 2) they mistakenly encode the dependent variable to be incidents of civil war, rather than as they claim the onset of a civil war, 3) they mistakenly calculate the predicted values for the alternative "non-theoretical" models that they compare to.  In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.

@@

@blair2020forecasting (B&S hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed. For the moment we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.  In short, they uphold their hypothesis that theory guided empirical research produces better conflict predictions than machine learning inspired efforts that are essentially ad hoc combinations of available variables. They arrive at this conclusion by examining the problem of predicting civil war onset. They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory including a \textit{kitchen sink} model with more than 1,000 covariates.

@@

1. How does the theoretically-driven escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
2. Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
3. How accurate were predictions using the escalation model for the first half of 2016?

To assess the first two questions, B&S use ICEWS data covering all major countries from 2001 to 2015. Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years. The main outcome variable is civil war onset, measured using Sambanis' civil war dataset.

Both the first and second questions above rely on comparing their escalation model to various alternative models. The same procedure is used in both cases:

1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
2. Estimate the escalation and other competing models. 
3. Create out-of-sample (OOS) predictions from each model using the test set. 
4. Calculate AUC-ROC measures for each set of OOS predictions.

To examine the first question, B&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:
- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other. These are directed, thus making for four total covariates.
- CAMEO: counts for all CAMEO event codes totalling $1,159$ covariates, which are mostly zero for any country in any month. 
- Average: unweighted average of the predictions from the four models briefly described above.

The corresponding results for each question are shown in B&S Tables 1 and 2, which we will examine further below. We accurately replicate their Tables 1 and 2, with very tiny differences. 
The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions. These robustness checks vary either (1) random forecast hyperparameter values, or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.

The second question, whether structural variables add to the escalation model, is assessed by comparing the original escalation model to four alternatives that incorporate annual, structural variables that are used in the PITF instability forecasting model:

@@

Finally, B&S used their escalation model to create forecasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed. This is summarized in B&S Table 3.

@@

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set $ for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.  Our belief is that these research decisions lead B&S to draw conclusion that are incorrect. The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.  We turn to discussing these five issues below.

@@

This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation. 
The reason to use a smoothed curve is that you have a continuous dependent variable (not the case here) which can produce any value of the dependent variable.  Wherein there is a discrete number of actual outcomes there is no justification for using smoothed ROC plots, nor for caculating statistics based on them.  Moreover, it is easy to compute the nonsmoothed ROC curve, and associated statistics, so computational ease does not justify this choice.

@@

ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data. In this case, there are also groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached. For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.  Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way. This ignores the fact that the actual ROC

Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC. The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models. It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel. In any case, the interpretation of results should not be sensitive to random variation, i.e. it should not depend on using a specific RNG seed. On the basis of these results, B&S conclude that the escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.

Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves. The Average model outperforms the escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.

As it turns out, this decision has a dramatic impact on the AUC-ROC calculations they use to support their empirical claims. As we show below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and this differentially impacts the models that are being compared. In fact, B&S's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC. These results are shown in Tables 1 and 2.  The Goldstein, CAMEO, and Average model are everywhere superior to the theoretically informed escalation model.  Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.

@@

We infer that the intent is that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. Their implementation, however, actually weighs the test set predictions using the PITF model predictions for the training set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly in the 6-month version of this file.] This appears to be a coding error.

@@

The implementation error aside, this split-population analog model is actually quite odd and does not actually replicate an analogue of the idea behind split-population modeling. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision (regression^[See further below. Although the RF models are used for a binary decision problem, the actual implementation uses regression RFs for continuous outcomes.]) trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does indeed have 200,000 decision trees. Furthermore, the PITF model predictions do not impact the way the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).

@@

@@

table2_nosmooth %>%
  dplyr::select(-Model, -horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of B\\&S Table 2: Test set AUC-ROC for escalation model with and without structural PITF contribution",
    label = "table2-fixed") %>%
  column_spec(2:6, width = "2cm") %>%
  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
  footnote("Differences from the original B\\&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
```

@@

This is not the case anymore. It outperforms in the 1-month version and under-performs the escalation model in the 6-month version.

@@

What impact did the ROC smoothing have overall on the results reported in B&S Tables 1 and 2? Tables \ref{tab:table1-benefit} and \ref{tab:table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC. Positive values indicate that smoothing increased a model's apparent performance. The escalation model is the only model that consistently had a benefit from smoothing. For all eight other models, smoothing sometimes gave a benefit, sometimes not.

A more dramatic difference stands out when we consider the overall average impact of smoothing across all flavors of a model reported in the columns of the tables. The Average, Goldstein, Quad, PITF Split Population, and Weighted by PITF models are slightly hurt by smoothing, but on the order of less than 0.01 in absolute magnitude. The "With PITF Predictors" model is hurt on the order of 0.02, i.e. it appears substantially worse with smoothing. The CAMEO and PITF Only models benefit slightly, on the order of 0.01 or less. The escalation only model on the other hand has an average boost of 0.058 to its AUC-ROC from smoothing. Considering the spread of AUC-ROC values if we compare across rows in B&S Tables 1 and 2, that boost is substantial.

The use of smoothed ROC curves to calculate AUC-ROC values only clearly benefits the escalation model, it does so consistently, and by a considerable margin. All eight alternative models reported in Tables 1 and 2 do not on average gain when using smoothed ROC curves to calculate AUC.

@@

- the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
- structural variables substantially improve the escalation model's performance when added to the pool of predictors on which the underlying random forest model is based.
